# Does sex bias exist at the university in the most current year available (1995)?

```{r echo = T, results = 'hide'}
library(dplyr)    # for data manipulation
library(ggplot2)  # for plotting
library(scales)   # for percentage scales in some plots (e.g. scale_y_continuous(labels=percent))
```


First, I will read in the whole dataset.
```{r}
salary_data <- read.table(file = "salary.txt", header = TRUE)
salary_data
```

Next, filter data to the year 1995.
```{r}
salary_data_1995 <- salary_data[salary_data$year == 95, ]
salary_data_1995
```

First, create visualizations for 1995 data: the distribution of academic field, year in which they attained their highest degree, year they were hired at the university, academic rank, and monthly salary in 1995 by sex.

```{r}
p_field_bar <- ggplot(salary_data_1995, aes(x = field, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(title = "Fields of Work by Sex in 1995 (Absolute Count)",
       x = "Field",
       y = "Count") +
  theme_minimal()
print(p_field_bar)
```
```{r}
salary_data_1995_prop <- salary_data_1995 %>%
  count(field, sex) %>%  # Count occurrences per field and sex
  group_by(field) %>%  
  mutate(prop = n / sum(n))  # Compute proportion

p_field_bar <- ggplot(salary_data_1995_prop, aes(x = field, y = prop, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +  # Use "stack" for a stacked bar chart
  labs(title = "Fields of Work by Sex in 1995 (Stacked Proportion)",
       x = "Field",
       y = "Proportion") +
  theme_minimal()

print(p_field_bar)
```

```{r}
salary_data_1995_prop <- salary_data_1995 %>%
  count(admin, sex) %>%  # Count occurrences per field and sex
  group_by(admin) %>%  
  mutate(prop = n / sum(n))  # Compute proportion

p_field_bar <- ggplot(salary_data_1995_prop, aes(x = admin, y = prop, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +  # Use "stack" for a stacked bar chart
  labs(title = "Possession of Admin Status by Sex in 1995 (Stacked Proportion)",
       x = "Admin",
       y = "Proportion") +
  theme_minimal()

print(p_field_bar)
```

```{r}
salary_data_1995_prop <- salary_data_1995 %>%
  count(deg, sex) %>%  # Count occurrences per field and sex
  group_by(deg) %>%  
  mutate(prop = n / sum(n))  # Compute proportion

p_field_bar <- ggplot(salary_data_1995_prop, aes(x = deg, y = prop, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +  # Use "stack" for a stacked bar chart
  labs(title = "Possession of Deg Status by Sex in 1995 (Stacked Proportion)",
       x = "Deg",
       y = "Proportion") +
  theme_minimal()

print(p_field_bar)
```



```{r}
yrdeg_means <- salary_data_1995 %>%
  group_by(sex) %>%
  summarise(mean_yrdeg = mean(yrdeg, na.rm = TRUE))

p7 <- ggplot(salary_data_1995, aes(yrdeg))
plot_of_sex_vs_yrdeg <- p7 + geom_density(aes(fill = factor(sex)), alpha = 0.8) + 
    geom_vline(
    data = yrdeg_means,
    aes(xintercept = mean_yrdeg),
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(title = "Sex vs Year when Highest Degree was attained in 1995 (Density Plot)", 
       x = "Year in which Highest Degree was Attained",
       fill = "sex")
plot_of_sex_vs_yrdeg
```

```{r}
hire_data <- salary_data_1995 %>%
  group_by(startyr, sex) %>%
  summarise(count = n(), .groups = 'drop')

avg_years <- hire_data %>%
  group_by(sex) %>%
  summarise(
    avg_year = sum(startyr * count) / sum(count),  # Weighted mean
    .groups = 'drop'
  )
avg_years
```

```{r}
p_hire_timeseries <- ggplot(hire_data, aes(x = startyr, y = count, color = sex)) +
  geom_line() +
  # 3. Add vertical lines for each sex's average hire year
  geom_vline(
    data = avg_years,
    aes(xintercept = avg_year, color = sex),
    linetype = "dashed",
    size = 1
  ) +
  labs(
    title = "Number of Hires By Start Year for Employees in 1995",
    x = "Year of Hire",
    y = "Count"
  ) +
  theme_minimal()
p_hire_timeseries
```

```{r}
p_hire_dens <- ggplot(salary_data_1995, aes(x = startyr, fill = sex)) +
  geom_density(alpha = 0.3, position = "fill") +
  labs(title = "Percentage of Hires by Start Year For Employees in 1995",
       x = "Year of Hire",
       y = "Proportion") +
  scale_y_continuous(labels = percent) +
  theme_minimal()
print(p_hire_dens)
```

```{r}
p_rank_95_count <- ggplot(salary_data_1995, aes(x = rank, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(title = "Academic Rank in 1995 - Total Counts",
       x = "Rank",
       y = "Count") +
  theme_minimal()
print(p_rank_95_count)
```

```{r}
p_rank_95_percent <- ggplot(salary_data_1995, aes(x = rank, fill = sex)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent) +
  labs(title = "Academic Rank in 1995 - Percentage by Sex",
       x = "Rank",
       y = "Percentage") +
  theme_minimal()
print(p_rank_95_percent)
```

```{r}
p_salary_95_box <- ggplot(salary_data_1995, aes(x = sex, y = salary, fill = sex)) +
  geom_boxplot() +
  labs(title = "Salary Distribution by Sex in 1995",
       x = "Sex",
       y = "Salary") +
  theme_minimal()
print(p_salary_95_box)
```

## NEXT: Use linear regression to answer the question, "Does sex bias exist at the university in 1995?"
First, I will use simple linear regression with mean salary as the response variable and sex as the predictor variable. Then, I will use a more complicated multivariate linear regression that factors in more variables.

Here is a simple linear regression with mean salary as the response variable and sex as the predictor.
```{r}
simple_model <- lm(salary ~ sex, data=salary_data_1995)
summary(simple_model)
```
As we can see, sex is a predictor of salary. If we let $x_1$ be an indicator for male sex and $\beta_1$ be its coefficient, we reject the null hypothesis that $\beta_1 = 0$ with an alpha value of $\alpha < 2e-16$.

Perhaps some of this difference can be explained by looking at other variables in the dataset. From the exploratory data analysis above, there appear to be discrepancies by sex with respect to rank, field, degree, year degree, start year, and admin. I suspect year degree and start year may be related, but I will begin by including both in my model.

```{r}
model2 <- lm(salary ~ sex + rank + field + deg + yrdeg + startyr + admin, data=salary_data_1995)
summary(model2)
```

Without yrdeg:
```{r}
model3 <- lm(salary ~ sex + rank + field + deg + startyr + admin, data=salary_data_1995)
summary(model3)
```

without startyr:
```{r}
model4 <- lm(salary ~ sex + rank + field + deg + yrdeg + admin, data=salary_data_1995)
summary(model4)
```

When we exclude yrdeg, we see that startyr's p-value is high, 0.303. We will remove it, as when only yrdeg is included it has a p-value of 1.34e-08.

```{r}
model5 <- lm(salary ~ sex + rank + field + deg + admin, data=salary_data_1995)
summary(model5)
```

